---
title: "Initial QC"
author: "Caroline_Duncombe"
date: "2024-08-07"
output: html_document
---

Goal:
Create GeoMX Seurat object and perform DE analysis
# Load in and normalize data
```{r load the libraies }
library(here)
library(GeomxTools)
library(CellMembrane)
library(DESeq2)
library(Seurat)
#load or download the 'usethis' package.
pacman::p_load("usethis")
pacman::p_load(BiocManager, devtools)
```

```{r Load Cell Membrane if necessary}
if (! ("pacman" %in% installed.packages()[,"Package"])) {
  install.packages('pacman')
}
#load or download the 'usethis' package.
pacman::p_load("usethis")


usethis::create_github_token()

gitcreds::gitcreds_set()

pacman::p_load(BiocManager, devtools)

# Make sure to update your Rprofile to include Bioconductor repos, such as adding this line to ~/.Rprofile:
local({options(repos = BiocManager::repositories())})

#Latest version:
devtools::install_github(repo = 'bimberlabinternal/cellmembrane', dependencies = TRUE, upgrade = 'always')


```

```{r Load the datafile}

download_directory <- here::here('raw_data/dcc_files/DCC-20240715')

# Point the readNanoStringGeoMxSet function to the directory containing DCC files from this plate
dccFiles <- dir(download_directory, pattern = "\\w*.dcc$", full.names = TRUE)
#dccFiles <- download_directory
  
  # Point to the pkc file (kind of like a reference atlas for nanostring)
  pkcFile <- "raw_data/pkc/Mm_R_NGS_WTA_v1.0.pkc"
  
  # Point to the annotation file (metadata for the samples)
  annotationFile <- "raw_data/meta_data/annotation_data (2).xlsx"
  
  myData <- suppressWarnings(readNanoStringGeoMxSet(dccFiles = dccFiles,
                                                    pkcFiles = pkcFile,
                                                    phenoDataFile = annotationFile,
                                                    phenoDataSheet = "Master",
                                                    # this is the excel sheet’s name
                                                    phenoDataDccColName = "SAMPLE_ID",
                                                    # these are headers/column names in the excel file
                                                    protocolDataColNames = c("ROI_ID") # add columns about protocol/sequencing data here. 
                                                    ))
  
  # Aggregate counts (for the WTA, this just aggregates negative probes)
  target_myData <- aggregateCounts(myData)
  
  # The data won't export unless a normalization method is run, but I only want the raw data
  # Using for housekeeping genes. 
  target_myData <- normalize(target_myData, norm_method = "hk", fromElt = "exprs", toElt = "hk_norm")
  
  # Export the raw data to a Seurat object
  seuratObj.tmp <- GeomxTools::as.Seurat(target_myData, normData = "exprs", forceRaw = TRUE)
  
  
  # playing with Seurat
  Seurat::GetAssayData(seuratObj.tmp, layer = "counts") # coudl add meta data to this. 
  
seuratObj.tmp@meta.data # this is our normal dataframte
  
```

```{r Normalize with CellMembrane and create Seurate object}

# Normalization with CellMemraben - there is the normalization process and there is not.
# Fit a model hat is sensitive to the things that you want to take into account. 
#This does a  library sized tranformation - transcript per million reads per library.
# Tihs is only transcripts that got aligned. 
# Seurat does per 100,000 log2 or log - see what seurat
# This is just to run and see a rough PCA. 

seuratObj.tmp <- CellMembrane::NormalizeAndScale(seuratObj.tmp, 
                                                       nVariableFeatures = 5000, 
                                                       scoreCellCycle = F)
 
seuratObj.tmp<- CellMembrane::RunPcaSteps(seuratObj.tmp, npcs = 50)

Seurat::DimPlot(seuratObj.tmp, reduction = 'pca', dims = c(1,2), group.by = "ROI_size")
Seurat::FeaturePlot(seuratObj.tmp, feature = "Cyp4a12a", reduction = 'pca', dims = c(1,2))

#sequester data into subsets
seuratObj.100 <- subset(seuratObj.tmp, subset = ROI_size == "100um")

Seurat::DimPlot(seuratObj.100, reduction = 'pca', dims = c(1,2), group.by = "ROI_size")

#####
seuratObj.tmp <- NormalizeData(seuratObj.tmp)
seuratObj.tmp <- FindVariableFeatures(seuratObj.tmp, nfeatures = 5000) #5000 is completely made up
seuratObj.tmp <- ScaleData(seuratObj.tmp)
 
seuratObj.tmp <- RunPCA(seuratObj.tmp)

#Could review the result of the bulk data - what is the same and the spatially informed liver. 
#What is the same versus different. 
  
```

```{r Lookt at Seurate features for PCA plots}

print(seuratObj.tmp@meta.data)

pbmc_small <- CellMembrane::FindClustersAndDimRedux(seuratObj.tmp)

Seurat::DimPlot(pbmc_small, group.by = 'Cyp4a12a')
Seurat::FeaturePlot(pbmc_small, features = 'Cyp4a12a')


```

# Next step: Preliminary evaluation of analysis models.

Goal is to data harvesting to set up Deseq2.
potential formulte: formula = 0 + Sex * Location # within infected

```{r DE  with an intercept model - all variables}
# need matrix and metadata:
cts <- Seurat::GetAssayData(seuratObj.tmp, layer = "counts")
coldata <- seuratObj.tmp@meta.data

#Intercept one - always compared against female
dds <- DESeqDataSetFromMatrix(countData = round(as.matrix(cts),0),
                             colData = coldata,
                             design= ~ 0 + Group + ROI_size + Liver_zone + Type)

#0 + Group*Liver_zone # fiting these and getting volcano plots
#0 + Liver_zone*Type # fiting these and getting volcano plots

dds <- DESeq(dds)
resultsNames(dds) # lists of comparions:
#"Intercept"                  "Group_M_vs_F"               "Group_ORX_vs_F"             "ROI_size_200um_vs_100um"   
 #[5] "ROI_size_50um_vs_100um"     "ROI_size_75um_vs_100um"     "Liver_zone_PC_vs_IZ"        "Liver_zone_PP_vs_IZ"       
 #[9] "Type_mock_vs_bystander"     "Type_schizont_vs_bystander"

```

```{r DE without the intercept - all variables}

dds2 <- DESeqDataSetFromMatrix(countData = round(as.matrix(cts),0),
                             colData = coldata,
                             design= ~ Group + ROI_size + Liver_zone + Type)
 #[1] "GroupF"        "GroupM"        "GroupORX"      "ROI_size200um" "ROI_size50um"  "ROI_size75um"  "Liver_zonePC"  "Liver_zonePP"  "Typemock"      "Typeschizont" 

dds2 <- DESeq(dds2)
resultsNames(dds2) # lists the groups:

```

``` {r Running the comparisons}
res2 <- results(dds2, name="Group_M_vs_F") 
res2 <- lfcShrink(dds2, coef= c("Group_M_vs_F"), type="apeglm")


res <- results(dds, name="GroupM") 
# or to shrink log fold changes association with condition:
res <- lfcShrink(dds, coef= c("GroupM"), type="apeglm")

# look at it:
res
# Change versus global mean
# Male versus female

```

Create volcano plots from some of the comparions

Function to make volcano plots (pulled from GW code and altered from edgeR to DeSeq2)
```{r Function to perform DE expression and volcano plot}
#' @title PerformDifferentialExpression
#'
#' @description This is designed to perform DE using edgeR on a pseudobulked seurat object
#' @param fit The glm fit object generated by PerformGlmFit
#' @param contrast The contrast, passed directly to edgeR::glmQLFTest or edgeR::glmLRT
#' @param contrast_name A name for the contrast
#' @param logFC_threshold The min logFC
#' @param FDR_threshold The min FDR to report
#' @param test.use Can be either QLF or LRT. QLF runs edgeR::glmQLFTest, while LRT runs edgeR::glmLRT
#' @param showPlots Boolean determining if the volcano plots and p value dsitributions should be shown. 
#' @return A list with the differential_expression results, volcano ggplot object and pvalue_dist ggplot object
#' @export
#' 
PerformDifferentialExpression <- function(dds, contrast_name, logFC_threshold = 1, FDR_threshold = 0.05, test.use = "Wald", showPlots = TRUE){

  # Perform differential expression analysis for the given contrast
  if (test.use == "Wald"){
    fit_test <- DESeq2::results(dds, name = contrast_name)
  } else if (test.use == "LRT"){
    fit_test <- DESeq2::results(dds, contrast = contrast_name, test = "LRT")
  } else {
    stop("Please supply a valid test.use argument.")
  }

  # Shrink log2 fold changes (optional, recommended for visualization)
  fit_test <- DESeq2::lfcShrink(dds, coef = contrast_name, type = "apeglm")
  
  # Convert results to a data frame and add a 'gene' column
  differential_expression <- as.data.frame(fit_test)
  differential_expression$gene <- rownames(differential_expression)
  
  # Create volcano plot
  volcano <- ggplot(differential_expression, aes(x = log2FoldChange, y = -log10(padj), label = gene)) +
    # Plot the upregulated genes using log2FoldChange cutoff and padj cutoff
    geom_point(data = differential_expression[differential_expression$log2FoldChange >= logFC_threshold &
                                              differential_expression$padj <= FDR_threshold,], color = "orange") +
    # Plot the downregulated genes
    geom_point(data = differential_expression[differential_expression$log2FoldChange <= -logFC_threshold &
                                              differential_expression$padj <= FDR_threshold,], color = "cadetblue2") +
    # Repel labels for the up- and down-regulated genes
    ggrepel::geom_text_repel(data = differential_expression[
      (differential_expression$log2FoldChange >= logFC_threshold & differential_expression$padj < FDR_threshold) |
        (differential_expression$log2FoldChange <= -logFC_threshold & differential_expression$padj < FDR_threshold),],
      max.overlaps = 10) +
    # Plot the rest of the genes
    geom_point(data = differential_expression[(differential_expression$log2FoldChange > -logFC_threshold &
                                               differential_expression$log2FoldChange < logFC_threshold) |
                                               differential_expression$padj > FDR_threshold,], color = "black") +
    # Plot thresholds
    geom_hline(yintercept = -log10(FDR_threshold), color = "red", linetype = "dashed") +
    geom_vline(xintercept = logFC_threshold, color = "red", linetype = "dashed") +
    geom_vline(xintercept = -logFC_threshold, color = "red", linetype = "dashed") +
    egg::theme_article() +
    ggtitle(contrast_name)
  
  # Create p-value distribution plot
  pvalue_dist <- ggplot(differential_expression, aes(x = pvalue)) + 
    geom_histogram() + 
    ggtitle("PValue Distribution") + 
    egg::theme_article()
  
  # Display plots if requested
  if (showPlots) {
    print(pvalue_dist)
    print(volcano)
  }
  
  # Return results
  return(list(
    differential_expression = differential_expression,
    volcano = volcano,
    pvalue_dist = pvalue_dist)
  )
}

```

```{r Create a Volcano plot with function}

PerformDifferentialExpression(dds, "Typemock") # Add comparison of interest. 

```

Create volcano plots based on Deseq outcomes

Follow GW McElfresh vignettes.

https://bimberlabinternal.github.io/CellMembrane/articles/Pseudobulking-and-Filtered-Contrasts.html
Pseudobulking-and-Filtered-Contrasts

https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#input-data

Checking for positive controls & interactions. 

### Summarize results from initial DE investigations with an upset plot. 
```{r GW to paste upset plot here}



```


