---
title: "QC evaluations of transcriptomic data"
author: "Caroline_Duncombe"
date: "2024-08-13"
output: html_document
---

# Overview of GeoMX sample omission process

There are several methods for filtering samples for GeoMX. There could be the Nanostring recommended approach, where using the point and click option you are filtering based on their 
* (1) technical signal criteria
* (2) biological probe QC process
* (3) filtering down of the genes (target) that have low signal and low variance. 

```{r load libraries}

library(here)
library(GeomxTools)

```

```{r Prepare GeoMx object}
download_directory <- here::here('raw_data/dcc_files/DCC-20240715')

  # Point the readNanoStringGeoMxSet function to the directory containing DCC files from this plate
dccFiles <- dir(download_directory, pattern = "\\w*.dcc$", full.names = TRUE)
#dccFiles <- download_directory
  
  # Point to the pkc file (kind of like a reference atlas for nanostring)
  pkcFile <- "raw_data/pkc/Mm_R_NGS_WTA_v1.0.pkc"
  
  # Point to the annotation file (metadata for the samples)
  annotationFile <- "raw_data/meta_data/annotation_data (2).xlsx"
  
  myData <- suppressWarnings(readNanoStringGeoMxSet(dccFiles = dccFiles,
                                                    pkcFiles = pkcFile,
                                                    phenoDataFile = annotationFile,
                                                    phenoDataSheet = "Master",
                                                    # this is the excel sheet’s name
                                                    phenoDataDccColName = "SAMPLE_ID",
                                                    # these are headers/column names in the excel file
                                                    protocolDataColNames = c("ROI_ID") # Add columns about sequencing data here. 
                                                    ))
  
```

After reading in the object, we will do a couple of QC steps.

Shift all 0 counts by 1
Flag low quality ROIs
Flag low quality probes
Remove low quality ROIs and probes

```{r}
demoData <- myData
#demoData <- shiftCountsOne(demoData, useDALogic=TRUE) # shifts all 0 counts in the file to 1 (why is this necessary?)
demoData <- setSegmentQCFlags(demoData, 
                  qcCutoffs=list(minSegmentReads=1000, 
                                 percentAligned=80, 
                                 percentStitched=80,
                                 percentTrimmed=80,
                                 percentSaturation=50,
                                 minNegativeCount=0, 
                                 maxNTCCount=1000,
                                 minArea=1800,
                                 minNuclei=2))

demoData <- setBioProbeQCFlags(demoData,
                    qcCutoffs=list(minProbeRatio=0.1,
                                   percentFailGrubbs=20),
                   removeLocalOutliers=FALSE)

demoflag <- protocolData(demoData)[["QCFlags"]]    # 5 samples failed
probeflag <- featureData(demoData)[["QCFlags"]]    # no probes failed

QCResultsIndex <- which(apply(protocolData(demoData)[["QCFlags"]], 
                              1L , function(x) sum(x) == 0L))
QCPassed <- demoData[, QCResultsIndex]
dim(QCPassed) 
# Features  Samples 
# 20175      273 

#
# Note: Samples in meta seem off from DCC file. How do I check this? The samples filtered don't match what I have found. 

#Decide whether to use the samples that did not pass in the downstream analysis.

```


# Process conducted with RAM

## Technical signal 
Read threshold
Want segments with less than 1000 raw reads

Aligned reads
80% raw read aligned
*If failed this nto great. Omit from analysis. 

Sequnecing Saturation - qdequately sequenced the library. Not confident you have captured. Might indicate you have seqeunced it enough. 
50%
*if not passed, need to also omit. 

Want to adjust the Nuclei Count
Surface are et minumum to 1800

## Technical signal 


Stiched - %
realted to the alignment of primers. 

## We have done the QC on the samples.
On the point and clide
Omited 5, did not exclude by nuclei count. 


# Biological Probe QC
leave on the default parameters.
Looks at if probes are consitent across ROIs. 
All the probes passed. 

Outcome- keeping all the probes in the dataset.

Coudl be more stringent - but do not do.

# segements = ROIS.
Don't filter ROIs at this stage.

# Step 3 - filter the target (genes) for each ROIs
This would this be donw in R.
High signal with low variablility.
Ram recommends the keeping the targets that 5%. Remvoing the highest variability and dn lowest number. 
Set to the Higher of LOQ and user defined value.

We set to 2% which leaves the about 10,000 genes. 

#Next step don't use the data normalization. 

Nanostring recommend the Q3 normalization strategy. looking at 75th quartile.
The low sized ones might be problemnation. Number of counts and the spread of them. 


# there are 19963
# you are not sequencing directly the RNA moelcule but sequecning the probe that was bound to the RNA molecule. 


# Now try to do this:

What we got:
* DCC files - Scott groups do seqeuncing and then they processing and created a DCC file for us. Input fastq and output DCC. 
* PKC file - from nanostring - connects probe ID to a target gene.
* fastq files - get from Scott Kennedy - do it now rather than forget. 
* High resolution image of the GeoMX images. 
* library readout - information sent on libraries to Scott Kennedy.
* a version of normalization. 

# Theo Bammler

Referred by Ram.
They do end-to-end analysis. 
3000 - 5000 - For end to end analysis. 
There might be better methods for normalization. 
